Think again?\\ Optimal mental simulation tracks problem difficulty

### 2015-02-01 manuscript.tex
Add keywords
Merge modeling sections

\textbf{Keywords:} 
add your choice of indexing terms or keywords; kindly use a
semicolon; between each term
\end{abstract}

@@

\section{The SPRT strategy}

We consider binary (or two-alternative forced choice) decisions, where an agent must choose one of two hypotheses, $H_0$ or $H_1$. 
The agent may take samples $X_i$ from a Bernoulli distribution parameterized by an unknown parameter $p$ (the probability of sampling evidence for $H_1$), and from these samples estimate the probability that $H_1$ is correct: $\hat{p}=\frac{1}{N}\sum_{i=1}^N X_i$, where $N$ is the total number of samples. 
Then, the decision rule which minimizes the probability of error is $\hat{H}(X_1,\ldots{},X_N)=H_0$ when $\hat{p}<0.5$ and $\hat{H}(X_1,\ldots{},X_N)=H_1$ when $\hat{p}>0.5$.

@@

\section{Experiment}

@@

@@

\section{Modeling mental sampling}

In the previous section, we presented an experiment in which participants had to judge whether a ball would go through a hole.
The results indicate a clear relationship between people's responses and response times---one which appears qualitatively similar to the one we would expect from SPRT.
In this section, we flesh out our hypothesis that people predict where the ball will go by running noisy physical simulations, and that they run a variable number of simulations on each trial based on SPRT principles \cite{wald1947sequential}.

\subsection{The physical simulator}

There is a growing body of evidence that people reason about physical scenes by running noisy simulations.
This hypothesis, referred to as the ``noisy Newton'' hypothesis, states that people have approximate knowledge of physical laws instantiated in a runnable model of intuitive physics.
Using this model, they can extrapolate the future of physical scenes by running a series of noisy simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}.
In particular, \citeA{Smith:2013fc} investigated the various sources of uncertainty in these simulations, finding that people's judgments were best captured by a model that took into account both perceptual uncertainty (noise in exactly where objects are and their motion trajectories) and dynamic uncertainty (stochastic noise added to account for unknowable variation in motion---e.g., a textured floor would cause a ball rolling along it to deviate from a straight line).

Specifically, \citeA{Smith:2013fc} asked people to catch a ball like the one in Figure \ref{fig:experiment} using a paddle that could move up and down along the $y$-axis.
In order to predict where people would place the paddle on each trial, their model incorporated uncertainty in its physical simulations through five parameters: perceptual uncertainty over the position of the ball ($\sigma_p$), uncertainty in the initial direction of velocity ($\kappa_v$), ongoing uncertainty in the direction of movement ($\kappa_m$), uncertainty in the bounce angle ($\kappa_b$), and a ``center bias'' reflecting a bias for people to respond more towards the center of the $y$-axis ($\sigma_0$).

Here, we used the model of \citeA{Smith:2013fc} to describe how people simulate individual possible physical events.
Because the hole-choice experiment was performed online, we performed an online replication of that experiment to set model parameters (see the Appendix for details).
Thus, the physical prediction parameters were set independently of the results of the experiment in this paper.

\subsection{The SPRT sampler}

\begin{figure}[t]
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/model_rt_results.pdf}
        \caption{\textbf{Model vs. human response times.} Each point corresponds to a different stimulus, trial type, and hole size; color and shape indicate the number of times the ball bounced during feedback. The $x$-axis is the fitted model response times, and the $y$-axis is the logarithmic mean response times of participants.}
        \label{fig:model-rt-results}
    \end{center}
\end{figure}

\subsection{Modeling reaction time}

@@

The fitted model explains participants' judgments of whether the ball would go in the whole very well (\HoleResponseCorr{}, see Figure \ref{fig:model-response-results}), and can additionally account for how quickly participants responded to trials (\HoleRTCorr{}, see Figure \ref{fig:model-rt-results}).
Even if the number of bounces is not included, the SPRT model with \threshold{} is able to account for a moderate amount of the variance in response times (\NoBouncesHoleRTCorr{}).
In contrast, a simpler model that takes only a single sample each time (equivalent to SPRT with $T=1$) does slightly better at explaining people's responses (\RawHoleResponseCorr{}) but cannot explain the variation in response times beyond the number of bounces.

@@

\begin{figure}[t]
    \begin{center}
        \includegraphics[width=0.45\textwidth]{figures/model_response_results.pdf}
        \caption{\textbf{Model vs. human judgments.} Each point corresponds to a different stimulus, trial type, and hole size. The $x$-axis is the probability the model says the ball will go in the hole, and the $y$-axis is the proportion of participants saying the ball will go in the hole.}
        \label{fig:model-response-results}
    \end{center}
\end{figure}

@@

@@

@@

@@

>>>>

\textbf{Keywords:} 
mental simulation; intuitive physics; SPRT; computational modeling
\end{abstract}

@@

\section{Making decisions from mental simulations}

Consider the task in Figure \ref{fig:experiment}, in which people observe a ball moving inside a box, and must predict whether it will go through a hole or not. How do people solve this problem?
In this section, we formalize a model that answers this question by combining noisy physical simulations with a decision-making strategy known as the \emph{sequential probability ratio test}, or SPRT.

\subsection{Modeling physical simulation}

There is a growing body of evidence that people reason about physical scenes like the one in Figure \ref{fig:experiment} by running noisy simulations.
This hypothesis, referred to as the ``noisy Newton'' hypothesis, states that people have approximate knowledge of physical laws instantiated in a runnable model of intuitive physics.
Using this model, they can extrapolate the future of physical scenes by running a series of noisy simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}.

\citeA{Smith:2013fc} investigated the various sources of uncertainty in these simulations, finding that people's judgments were best captured by a model that took into account both perceptual uncertainty (noise in exactly where objects are and their motion trajectories) and dynamic uncertainty (stochastic noise added to account for unknowable variation in motion---e.g., a textured floor would cause a ball rolling along it to deviate from a straight line).
Specifically, \citeA{Smith:2013fc} asked people to catch a ball like the one in Figure \ref{fig:experiment} using a paddle that could move up and down along the $y$-axis.
In order to predict where people would place the paddle on each trial, their model incorporated uncertainty in its physical simulations through five parameters: perceptual uncertainty over the position of the ball ($\sigma_p$), uncertainty in the initial direction of velocity ($\kappa_v$), ongoing uncertainty in the direction of movement ($\kappa_m$), uncertainty in the bounce angle ($\kappa_b$), and a ``center bias'' reflecting a bias for people to respond more towards the center of the $y$-axis ($\sigma_0$).

Here, we used the model of \citeA{Smith:2013fc} to describe how people simulate individual possible physical events.
Because the experiment in Figure \ref{fig:experiment} (see Experiment) was performed online, we additionally performed an online replication of the experiment from \citeA{Smith:2013fc} to set model parameters (see the Appendix for details).
Thus, the physical prediction parameters were set independently of the results of the experiment in this paper.

\subsection{The SPRT strategy}

If people are running simulations to reason about physical scenes, then a natural question is: how many simulations do they run?
Because the simulations are noisy, it might be benefial to run multiple simulations in order to get a better estimate of the outcome.
However, each simulation comes with a time cost.
To optimize this speed-accuracy tradeoff, we apply the \emph{sequential probability ratio test}, or SPRT \cite{wald1947sequential} to the samples drawn from the physical simulation.
By combining these two models, we can make predictions both for people's judgments, and how long they take to make those judgments.

Formally, we consider binary (or two-alternative forced choice) decisions, where an agent must choose one of two hypotheses, $H_0$ or $H_1$. 
The agent may take samples $X_i$ from a Bernoulli distribution parameterized by an unknown parameter $p$ (the probability of sampling evidence for $H_1$), and from these samples estimate the probability that $H_1$ is correct: $\hat{p}=\frac{1}{N}\sum_{i=1}^N X_i$, where $N$ is the total number of samples. 
Then, the decision rule which minimizes the probability of error is $\hat{H}(X_1,\ldots{},X_N)=H_0$ when $\hat{p}<0.5$ and $\hat{H}(X_1,\ldots{},X_N)=H_1$ when $\hat{p}>0.5$.

@@

\section{Experiment: Will the ball go in the hole?}

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=\textwidth]{figures/model_results.pdf}
        \caption{\textbf{Model vs. human comparison.} In both plots, each point corresponds to a different stimulus, trial type, and hole size. Dashed lines indicate perfect correspondence between model and people. \emph{Left}: The $x$-axis is the probability the model says the ball will go in the hole, and the $y$-axis is the proportion of participants saying the ball will go in the hole. \emph{Right}: Color and shape indicate the number of times the ball bounced during feedback. The $x$-axis is the model response times, and the $y$-axis is the logarithmic mean response times of participants.}
        \label{fig:model-results}
    \end{center}
\end{figure*}

@@

\subsubsection{Model comparison}

@@

The fitted model explains participants' judgments of whether the ball would go in the whole very well (\HoleResponseCorr{}, see Figure \ref{fig:model-results}A), and can additionally account for how quickly participants responded to trials (\HoleRTCorr{}, see Figure \ref{fig:model-results}B).
Even if the number of bounces is not included, the SPRT model with \threshold{} is able to account for a moderate amount of the variance in response times (\NoBouncesHoleRTCorr{}).
In contrast, a simpler model that takes only a single sample each time (equivalent to SPRT with $T=1$) does slightly better at explaining people's responses (\RawHoleResponseCorr{}) but cannot explain the variation in response times beyond the number of bounces.

@@

@@

\subsubsection{Participants}

@@

\subsubsection{Stimuli}

@@

\subsubsection{Procedure}

@@

\subsubsection{Results}
